---
title: "Formal coverage"
author: "Hugh Parsonage"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FormalCoverage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitrOpts}
library(knitr)
pkgs_avbl <-
  requireNamespace("testthat", quietly = TRUE) &&
  requireNamespace("covr", quietly = TRUE)
knitr::opts_chunk$set(eval = pkgs_avbl)
```

## Testing
R package development is supported by a mature ecosystem of diagnostic tools to prevent bugs.
Unit tests are available in base R and the package `testthat` by Hadley Wickham is commonly used.
A unit test ensure that a function with a given set of inputs returns a known result and ensures this known result survive any changes to the function's code.

### `testthat` and `covr`

```{r test-that}
add_numbers <- function(x, y) {
  x + y
}

library(testthat)
expect_equal(add_numbers(1, 1), 2)

```

However, unit tests do not ensure that *all* user inputs will result in correct code. Code coverage tools such as through the `covr` package complement unit tests by identifying any untested source code. 

```{r code-cov}
divide_numbers <- function(x, y) {
  if (y == 0) {
    stop("y cannot be zero.")
  }
  x / y
}

expect_equal(divide_numbers(12, 4), 3)

library(covr)
function_coverage(divide_numbers, 
                  code = expect_equal(divide_numbers(12, 4), 3))
```

Yet even with unit testing satisfying 100\% code coverage, bugs may still occur. 

```{r add3}
add3 <- function(x, y, z = 0) {
  x + y
}

expect_equal(add3(1, 1), 2)
function_coverage(add3, 
                  code = expect_equal(add3(1, 1), 2))
```

Despite having a unit test and 100\% coverage, the above function has a serious bug. 
No testing suite can eliminate all bugs, the goal of this package is to identify a narrow set of bugs wherein a function has an argument available to the user but which is never invoked in the body of the argument.
Whereas code coverage is used to tell whether all the lines in the *body* have been used in a test, `formal coverage' is used to tell whether all the *arguments* (or *formals*) of a function have been used in a test.

### `unused_formals` 

The function `unused_formals` takes a function and checks whether each argument is present in the body of the function.
If any arguments are not present but ought to be, those arguments are returned.
```{r unused_formals}
library(formalCoverage)
unused_formals(add3)
```

You can also supply a quoted version of the function, and specify the namespace:

```{r unused_formals_testthat}
unused_formals("expect_equal", ns = "testthat")
```

If all arguments are indeed used, the function returns `NULL` invisibly. 

Lastly, a function `formalPackageCoverage()` is available. 
It is run on the top level of your package directory, returning `NULL` if all the *exported* functions have formal coverage, or a named list of the offending functions and arguments otherwise. 

## Limitations
This package is immature and has limitations.
In that spirit, the package is quite shy: if there is any doubt, the test will succeed.

In particular, if the function's body contains `UseMethod`, the function is assumed to be covered.




